The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-28 manuscript/rebuttal.Rmd

RESPONSE: We thank reviewer 1 for their positive feedback.
With regards to the ordering of the five pillars, we have amended Figure 1 so that the pillars now
appear in the same order as they are detailed in sections following.

## Reviewer 2

> Comments to the Author: The article offers a comprehensive perspective on reproducibility in
> computational research, delivering clear and logically backed recommendations. It is indeed a very
> interesting topic that needs our attention. Nevertheless, there are areas that could benefit from
> further development and clarity:

RESPONSE: We appreciate reviewer 2's positive feedback, insights and thoughtful suggestions.

@@

### R2 Point 1

> 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case
> studies or practical examples that demonstrate the effectiveness of your recommendations.
> Highlighting specific instances where these best practices have enhanced reproducibility or solved
> particular issues could substantiate your points more effectively.

RESPONSE: To demonstrate the importance of computational reprodcibility, we elaborate on the
"forensic bioinformatics" case study of Baggerly and Coombes (2010)@Baggerly2010-nr critique of Potti
et al (2006) @Potti2006-yk in the introduction:

@@

### R2 Point 2

> 2. Expanding on Current Shortcomings: The article mentions that current practices are not meeting the
> goal of reproducibility. An expanded discussion on the ramifications of these shortcomings on
> scientific progress could accentuate the urgency and significance of addressing this issue.

RESPONSE: In the introduction we have spoken to the consequences of poor computational research, and
emphasize the urgency of this issue (which complements the case study summary above):

@@

> The ramifications of irreproducible and unreliable research include misleading the community,
> wasting research funds, slowing scientific progress, eroding public confidence in science and
> tarnishing the reputation of associated institutions and colleagues.
> In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.

### R2 Point 3

> 3. Focus on Implementation Challenges: While the article acknowledges the importance of certain
> practices, an increased focus on potential obstacles researchers may face when trying to implement
> these suggestions, as well as strategies to overcome these challenges, would be insightful.

RESPONSE: We have re-written the Challenges section to highlight some of the barriers slowing the
adoption of best practices, with reference to previous contributions on the topic.
These are structured around what different stakeholders can do to improve the situation (including
funders, journals, institutions and individual researchers).
We specifically note the effect of incentives on motivation, and the role of automation for
routine reproducibility checking.

@@

### R2 Point 4

> 4. Clear Structure: Consider organizing the article into clearly defined sections such as
> 'Introduction', 'Background', 'Recommendations', 'Challenges', and 'Conclusion'.
> This could significantly enhance readability and comprehensibility.

RESPONSE: The subheading structure has been amended accordingly.

@@

## Reviewer 3

### R3 Point 1

> The paper is mostly well written and addresses an important topic: computational reproducibility and
> tools to achieve it. The authors describe five "pillars" under which various tools and techniques can
> be categorized. The paper is pretty comprehensive in what it covers and has useful supplementary
> material. My main issue with the paper is that it reiterates many things that have been covered in
> other papers. The authors cite 11 such papers. Many of the topics covered in submitted paper have
> already been covered well in other papers. I am most familiar with reference 13, which covers many of
> the same topics, although this paper goes into more detail on many issues. The authors could do more
> to differentiate their paper from previous ones and perhaps remove some topics that are already
> covered well elsewhere.

RESPONSE: We thank reviewer 3 for their careful assessment of our article and appreciate the suggestions
which have improved the manuscript.
Indeed, this article does reiterate a great deal of what has already been said, it is a review after
all.
We note that each of those recommendation articles mentioned have different emphases.
For example, some of them don't speak to the benefits of literate programming,
while others do not mention the importance of documentation.
Some are focused on specific concepts like infrastructure, containerisation or
continuous analysis.
Some are proof of principle workflows, which present a rigid combination of tools.
Here, we describe the ways computational researchers can make their own work more reproducible.
We are assuming that the readers are not advanced analysts, so some topics like pipelining systems
are covered in less detail and we refer the reader to those relevant reviews.
What does make this article unique is the five pillars infographic, and specific combination of
recommendations that incorporates new material (eg: Software Heritage, Guix, Continuous Analysis).
Our article also provides a handy list of recommended tutorials as a supplementary resource.

@@

### R3 Point 6

> 6. Page 3, Line 56: It is not necessarily true that analyses performed using web tools are not
> reproducible. Although rare, some web tools facilitate reproducibility by providing code or
> configuration files and/or allowing the apps to be executed locally.

RESPONSE: Thank you for raising this point. We have amended the section in question:

> While webtools are valuable for data exploration, there are worries that they undermine
> reproducibility for the sake of convenience [@Wijesooriya2022-ly].
> Transferring data between compute platforms is also discouraged.
> For example, having workflows that involve combinations of web-based and scripted tools require
> data transfer steps which are inefficient and error-prone.
> On the other hand, some web based tools excel at reproducibility.
> The web-based analysis platforms Galaxy and GenePattern enable sophisticated point-and-click
> bioinformatics analysis in the browser, and those workflows can also be shared in a reproducible way
> [@The_Galaxy_Community2022-dt; @Reich2006-na].
> Some webtools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
> or by allowing apps to be executed locally (e.g.: ShinyGO @Ge2020-qe).

@@

### R3 Point 11

> 11. What about computationally intensive tasks that must be performed using specialized computing
> environments like the cloud or clusters?

RESPONSE: We have amended the relevant section to suggest a specific workflow manager for compute
intensive applications.

> Projects involving computationally intensive tasks would benefit from a build/workflow automation
> solution.
> Guides for selecting and using such systems have been already reviewed by others [@Leipzig2016-bf; @Reiter2021-lh].
> But it is worth mentioning some of the most used solutions in bioinformatics, which include
> `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021], `CWL` [@Crusoe2022-cb],
> `WDL` [@Voss2017-vd] and `nextflow` [@Di_Tommaso2017-yo].
> The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
> for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
> need to be repeated.
> After fixing the issue, re-running the analysis would pick up at step 8 again,
> which saves labor and compute time.

@@

> 13. Section on version control: This section focuses mostly on using VC for software development
> (that is my interpretation).
> To be consistent with the introduction, it should focus more on data analyses.
> Although I use VC for analyses, I feel that simpler approaches are better in many cases.
> For example Dropbox and Google Drive provide some version-control and backup functionality and do
> not require the same level of knowledge as git.

RESPONSE: In accordance with this suggestion, we have rephrased some of the text to highlight the
utility of DVCS for data analysts.

> Although DVCSs come from the world of software engineering, data analysts can significantly
> benefit from their use in the life sciences [@ram2013; @blischak2016; @perez-riverol2016].

The bullet points following this passage outline the benefits of using DVCS.
While cloud storage like dropbox could protect against code loss, the other five points cannot
be satisfied.

@@

### R3 Point 14

> 14. Page 7: Many data analysts will not know what JupyterLab or VS Code are.
References are also needed.

RESPONSE: The phrase has been updated accordingly:

> `git` is typically used at the command line, however it is also incorporated into integrated
> development environments commonly used in bioinformatics including Rstudio, JupyterLab
> [@Kluyver2016jupyter] and VS Code [@noauthor_undated-fn].

@@

> While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
> [@Kurtzer2017-zq], Docker remains the most widely used containerisation system (according to GitHub
> stars).

### R3 Point 17

> 17. The Biocontainers project should be mentioned.

RESPONSE: Biocontainers is mentioned in the last paragraph of the "Compute environment control" section.

@@

### R3 Point 18

> 18. Page 10, lines 51-52: I disagree that the risk is small.
> There are many instances of using genomic data to identify individuals who have committed crimes.

RESPONSE: In light of this, we have changed the section in question:

> While deidentification of data has been widely used to safeguard participant privacy, this may not be
> suitable for genomics data due to the availability of existing public genetic data sets that heighten
> reidentification risks [@Heeney2011-ri].
> For example, the surnames of some male participants can be inferred based on the Y chromosome
> variants seen in deidentified genomic data together with public geneology web sites [@Gymrek2013-fz].
> To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
> Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
> frameworks designed to protect personal data in a way that perserves reproducibility [@Rehm2021-kt;@Thorogood2021-kt].

@@

> 19. Figures 2 and 4 are very similar to figures used in reference 13.

Figure 4 is adapted from the diagram on the Docker official website, which was originally published in
2015 or earlier (https://web.archive.org/web/20151023003338/https://www.docker.com/what-docker).
We have cited the original source.

Figure 2 is somewhat different from reference 13, as it shows an Rstudio window with windows for raw code
in text format beside the output document.
The figure in reference 13 shows a Jupyter notebook window, which has a different arrangement of code
and outputs.

We do not believe any changes to Figure 2 or 4 are necessary.

@@

RESPONSE: We acknowledge that it can sometimes be tiresome to reiterate things like the FAIR principles.
However summarising the key points succinctly will be in the interests of readers with limited time.
We note that the original description of the FAIR principles by Wilkinson and colleagues is ~5,000 words,
while this entire manuscript is ~6,500. We note that other reviewers did not insist on removing this
section.

### R3 Point 21

> 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR.
There are lots of problems with FAIRness in these repositories.

RESPONSE: While there may be problems with FAIRness with NCBI GEO and SRA, these repositries mostly
satisfy FAIR principles (eg: @Bonaretti2019-zs).
From the perspective of a researcher sharing genomic data, they are "good enough" as compared to the
well known alternatives (eg: general purpose repository, institutional data storage, supplementary file,
dropbox, on request, etc.).

@@

> There are no standards for organizing supplementary data both within and across journals
> and such data are often not readily discoverable or openly accessible
> (to those without a relevant journal subscription, for example).

They also cite an article by Caetano and Aisenberg that speaks directly to the findability and
accessibility @Caetano2014-bi:

> Many journals provide the option for authors to
> include supplementary data that are made available in the digital version of the publication.
> However, repositories increase data discoverability, are more reliable for long-term storage and
> make it possible to make data sets available under open access even
> when authors transfer copyright ownership to publishers.

In light of this argument, we have added citations to Caetano and Aisenberg in the updated text.

### 2023-08-28 manuscript/5pillars.Rmd
process validation

text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
text( 5 , 0.2 , lab="End-to-end automated process",cex=1.2)
text(5,7,lab="Process validation & code testing",cex=1.2)

@@

A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A set of files under version control is called a "repository" and represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed version control systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs (Subversion, Git, mercurial, etc), but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

@@

In addition, the `README` should outline the contents of the repository, how users can contribute to
the codebase and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.
A recently described set of recommendations for making software FAIR describes further measures
that data analysts could use to enhance their code documentation.

@@

When these principles are not integrated correctly and checked, problems readily emerge
[@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
This is why regular code testing is another recommended best practice, especially after making updates
to the code or data [@Rule2019-eh; @The_Turing_Way_Community2019-ig].
It is worthwhile to conduct such testing on a separate computer to understand how someone else in the
scientific community would fare reproducing such a workflow.

Moreover, some have suggested adapting software development
principles of unit tests and continuous integration/continuous delivery.
This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
made to a project's code base [@Beaulieu-Jones2017-vj; @Blischak2019-dq].
If the workflow executes successfully and tests are passed, the repository can be decorated with a
"badge" to inform users of the lack of errors in the code.
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves.
Continuous analysis might therefore be a future pediment (crowning feature) to the five-pillared
"temple" of reproducible research.

>>>>

text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
text( 5 , 0.2 , lab="End-to-end automated process",cex=1.2)
text(5,7,lab="Process validation & testing",cex=1.2)

@@

In bioinformatics, sharing code is becoming standard practice for reproducibility and transparency,
and is a requirement for many specialized journals [@Cadwallader2022-dq].
Code sharing appears to improve the rate of article citations [@Maitner2023-pv].
One of the most popular ways to share research code is through online software repositories
[@Cadwallader2022-od] with integrated version control.
A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A set of files under version control is called a "repository" and represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed version control systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs (Subversion, Git, mercurial, etc), but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

@@

In addition, the `README` should outline the contents of the repository, how users can contribute to
the codebase and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.
A recently described set of recommendations for making software FAIR describes further measures
that data analysts could use to enhance their code documentation @Patel2023-qr.

@@

When these principles are not integrated correctly and checked, problems can readily emerge
[@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
This is why regular code testing after making updates to the code or data is considered best practice
[@Rule2019-eh; @The_Turing_Way_Community2019-ig; @Noor2022-uc].
For bioinformatics software, testing may involve a suite of unit tests of each function, and another
integration test whether the functions are working properly together [@Sarma2016-ol].

For data analysts, the situation is somewhat different as a typical bioinformatics workflow might
involve hundreds of lines of code, so conducting tests at each step in the process would be impractical.
However incorporating such tests at key points in a workflow such as during quality control of input
data, after data cleaning, before statistical analysis and a summary of the findings.
While `if/else` statements can be used for testing, it is recommended to use a dedicated testing
package such as `testthat` for R [@Wickham2011-tt] or `pytest` for Python [@pytest7.4].
In the example in Box 1, `testthat` raises an error to halt the script if the test fails, while
the `if/else` statement requires an extra line to do this.

```{r,test,class.source = 'fold-show'}

data(iris)
numspecies <- length(unique(iris$Species))

# ifelse test
if ( numspecies == 3) {
  print("Iris data describes three species")
} else {
  stop("Iris data doesn't describe three species")
}

# testthat test
library(testthat)
data(iris)
testthat::expect_equal(numspecies, 3)

*Box 1: An example of a test in R using an if/else statement and with the `testthat` package.
The test is checking that the `iris` dataset describes three species.*

Literate programming also provides an opportunity to record sanity checks in an output report.
In R, commands like `dim()`, `length()`, `head()`, `str()` are a simple way to show the features of
a dataset to validate that data transformation steps are working as expected.

Successful execution of all literate scripts in a project is an important validation step
that proves that the scripts are free from code errors.
It is worthwhile to conduct such testing on a separate computer to understand how others would fare
reproducing such a workflow.
Testing extends to documentation as well, so it is important to update it so it remains in accordance
with the codebase, and actively seek feedback from others on its suitability.

In software development, continuous integration/continuous delivery (CI/CD) is a process of automated
software compilation and testing.
A derivation of this for data analysis has been termed "continuous analysis", and involves automatic
execution and testing whenever a change is made to a project's code base
[@Beaulieu-Jones2017-vj; @Blischak2019-dq].
If the workflow executes successfully and tests are passed, the repository can be decorated with a
"badge" to inform users of the lack of errors in the code.
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves.
Continuous analysis might therefore be a future pediment (crowning feature) to the five-pillared
"temple" of reproducible research.
